---
title: "2023 MAY 19 - Tornadoes in the US"
date: 2023-05-19
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

------------------------------------------------------------------------

The goal of this work is to apply my understanding of data towards real world projects using R and the tidyverse. The data used here is from the TidyTuesday project.

## Check Out This Week's Data

```{r Load , message=FALSE}
# load R packages for analysis
library(tidyverse)
library(tidytuesdayR)
```

```{r, message=FALSE}
# Download the weekly data and make available in the tt object.
# Using the last_tuesday function gives us the latest TidyTuesday data from today's date
tt <- last_tuesday("2023-05-19")
```

```{r, message=FALSE}
# Check out the available data
tt
```

This week's data comes from the [Severe Weather Maps, Graphics, and Data Page](https://www.spc.noaa.gov/wcm/#data) of the NOAA's National Weather Service Storm Prediction Center.

The following are the included variables:

| **Variable** | **Class** | **Description**                                                                                                                                                                      |
|--------------|-----------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| om           | integer   | Tornado number. Effectively an ID for this tornado in this year.                                                                                                                     |
| yr           | integer   | Year, 1950-2022.                                                                                                                                                                     |
| mo           | integer   | Month, 1-12.                                                                                                                                                                         |
| dy           | integer   | Day of the month, 1-31.                                                                                                                                                              |
| date         | date      | Date.                                                                                                                                                                                |
| time         | time      | Time.                                                                                                                                                                                |
| tz           | character | [Canonical tz database timezone](https://en.wikipedia.org/wiki/List_of_tz_database_time_zones).                                                                                      |
| datetime_utc | datetime  | Date and time normalized to UTC.                                                                                                                                                     |
| st           | character | Two-letter postal abbreviation for the state (DC = Washington, DC; PR = Puerto Rico; VI = Virgin Islands).                                                                           |
| stf          | integer   | State FIPS (Federal Information Processing Standards) number.                                                                                                                        |
| mag          | integer   | Magnitude on the F scale (EF beginning in 2007). Some of these values are estimated (see fc).                                                                                        |
| inj          | integer   | Number of injuries. When summing for state totals, use sn == 1 (see below).                                                                                                          |
| fat          | integer   | Number of fatalities. When summing for state totals, use sn == 1 (see below).                                                                                                        |
| loss         | double    | Estimated property loss information in dollars. Prior to 1996, values were grouped into ranges. The reported number for such years is the maximum of its range.                      |
| slat         | double    | Starting latitude in decimal degrees.                                                                                                                                                |
| slon         | double    | Starting longitude in decimal degrees.                                                                                                                                               |
| elat         | double    | Ending latitude in decimal degrees.                                                                                                                                                  |
| elon         | double    | Ending longitude in decimal degrees.                                                                                                                                                 |
| len          | double    | Length in miles.                                                                                                                                                                     |
| wid          | double    | Width in yards.                                                                                                                                                                      |
| ns           | integer   | Number of states affected by this tornado. 1, 2, or 3.                                                                                                                               |
| sn           | integer   | State number for this row. 1 means the row contains the entire track information for this state, 0 means there is at least one more entry for this state for this tornado (om + yr). |
| f1           | integer   | FIPS code for the 1st county.                                                                                                                                                        |
| f2           | integer   | FIPS code for the 2nd county.                                                                                                                                                        |
| f3           | integer   | FIPS code for the 3rd county.                                                                                                                                                        |
| f4           | integer   | FIPS code for the 4th county.                                                                                                                                                        |
| fc           | logical   | Was the mag column estimated?                                                                                                                                                        |

## Initial Thoughts and Questions

Based on the variables provided, there are likely a few interesting relationships to explore with regards to recorded tornadoes, including magnitudes, number of states affected, estimated property loss, and number of injuries and fatalities. Furthermore, because climate change is a pressing concern for many, how has the severity of tornadoes changed over the years?

## Data Wrangling and Exploratory Data Analysis

Let's create a series of smaller data frames to explore whether these variables have some sort of relationship that we can further explore

```{r Wrangle}
#
df1

#
df2

#
df3

#
df4

#
df5

```

## Visualize

Using your processed dataset, create your unique visualization.

```{r Visualize}
#

  
```

## Save Image

Save your image for sharing. Be sure to use the `#TidyTuesday` hashtag in your post on twitter!

```{r}

# This will save your most recent plot
ggsave(
  filename = "My TidyTuesday Plot.png",
  device = "png")

```

## Future Work

### Hypothesis:

There is a statistically significant

```{r}

```
