---
title: "TidyModels Screencast - Horror Movie Descriptions"
author: "Eugene Zhou"
date: "2023-05-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

This document is an adaptation of the ["Use OpenAI text embeddings for horror movie descriptions"](https://youtu.be/UsaZV8ROMSc) by Julia Silge.

The goal is to apply my training in R towards real world projects, including the use of OpenAI towards a data science project. The data used here is from the TidyTuesday project.

```{r packages, results='hide', message=FALSE}
#load r packages 
library(tidytuesdayR)
library(tidymodels)
library(httr)
```

## Data Exploration

```{r horror movie data, results='hide', message=FALSE}
#get horror movie data from GitHub using tidytuesdayR package
tuesdata <- tt_load("2022-11-01")
horror_movies <- tuesdata$horror_movies
```

```{r}
#create a sample of the horror movie data by filtering movies that are in the English language and remove movies that do not have an overview

#use slice sample to randomly select rows
set.seed(123)

horror_movies_df <- horror_movies %>%
  filter(!is.na(overview), original_language == "en") %>%
  slice_sample(n = 1000)

glimpse(horror_movies_df)
```

```{r}
#check out 3 random samples of movie overviews
sample(horror_movies_df$overview, size = 3)
```

Set OpenAI API keys as envinroment variables using `Sys.setenv()` function

```{r}
#make API call to OpenAI

#text embeddings are representations of text learned from large datasets

embeddings_url <- "https://api.openai.com/v1/embeddings"
auth <- add_headers(Authorization = paste("Bearer", 
                                          Sys.getenv("OPENAI_API_KEY")))

body <- list(model = "text-embedding-ada-002", input = horror_movies_df$overview)

```

```{r}
#call to OpenAI

resp <- POST(
  embeddings_url, 
  auth, 
  body = body, 
  encode = "json"
)

```

Check `resp$status_code` to confirm API call

```{r}
#200 = success
#401 = lack of valid auth credentials
resp$status_code
```

```{r}
#convert response from one large text chunk to a list using flatten

embeddings <- content(resp, as = "text", encoding = "UTF-8") %>%
  jsonlite::fromJSON(flatten = TRUE)

#pluck data 
#text is clustered by closeness
embed_extract <- embeddings %>% 
  pluck("data", "embedding")
```

```{r}
#add extracted embeddings as a new column in the horror movie data
horror_embed <- horror_movies_df %>%
  mutate(embeddings = embed_extract)

horror_embed %>%
  select(id, original_title, embeddings)
```

Create a matrix where every row is a movie and every column is an OpenAI embedding

```{r}
embeddings_mat <- matrix(
  unlist(horror_embed$embeddings), 
  ncol = 1536, 
  byrow = TRUE
)
```

## Similarity

```{r}
#compute a cosine similarity matrix

embeddings_sim <- embeddings_mat / sqrt(rowSums(embeddings_mat * embeddings_mat))
embeddings_sim <- embeddings_sim %*% t(embeddings_mat)
dim(embeddings_sim)
```

```{r}
horror_movies_df %>%
  slice(4) %>%
  select(title, overview)
```

```{r}
#check out 

enframe(embeddings_sim[4, ], name = "movie", value = "similarity") %>%
  arrange(-similarity)
```

```{r}
horror_movies_df %>%
  slice(c(935, 379, 380)) %>%
  select(title, overview)
```

## PCA

PC1 explains the most "variation" in the text

```{r}
#identify first 32 principal components
#use prcopm_irlba for faster computation

set.seed(234)
horror_pca <- irlba::prcomp_irlba(embeddings_mat, n = 32)
```

```{r}
augmented_pca <- as_tibble(horror_pca$x) %>%
  bind_cols(horror_movies_df)
```

### PC Plots

Plot principal components

```{r pressure}
augmented_pca %>%
  #PC1 vs PC2 colored by vote average
  ggplot(aes(x = PC1, y = PC2, color = vote_average)) + 
  geom_point(size = 1.2, alpha = 0.8) + 
  scale_color_viridis_c()
```

We observe that vote is unrelated to text. Movie description does not influence ratings.

```{r}
#make plot interactive
library(plotly)

#assign previously constructed plot to variable p
p <- augmented_pca %>%
  #PC1 vs PC2 colored by vote average
  ggplot(aes(x = PC1, y = PC2, color = vote_average)) + 
  geom_point(size = 1.2, alpha = 0.8) + 
  scale_color_viridis_c()

#run in ggplotly function to explore
ggplotly(p, tooltip = "text")
```
